The Gamma: Data Exploration through Iterative Prompting

### 2020-09-12 wip-design/paper.tex
Providers, merge design and motivation

\begin{figure}
\includegraphics[width=.54\columnwidth]{figures/thegamma-annot}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\end{figure}

@@

\begin{itemize}
\item We introduce the iterative prompting principle in The Gamma (Section~\ref{sec:overview})
  and show that it can be used for querying of distinct data sources including data tables, graph
  databases and data cubes (Section~\ref{sec:implementation}).

@@

\subsection{Providers}
\label{sec:overview-tps}

all the important logic

define DSLs

\subsubsection*{Data Cube Type Provider}
Our first type provider allows users explore data form a data cube. For example, the World
Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
countries in the world each year. The data set is a three-dimensional cube with dimensions
corresponding to countries, indicators and years. The following example, also shown in
Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:

\begin{thegamma}
worldbank.byCountry.'United States'
  .'Climate Change'.'CO2 emissions (kt)'
\end{thegamma}

As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
we restrict the cube to a two-dimensional plane. We then choose an indicator category
\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
years as keys and emission data as values. The World Bank provider also supports filtering
by year and indicator.

Another type provider available in The Gamma that is based on the data cube structure allows
the user to explore UK government expenditure:

@@

The dimensions of the cube are government services, years and value type (adjusted, nominal,
per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type. Our implementation
does not yet use a standardized data cube storage and so adding another data cube source
currently requires implementing a new type provider.

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the  % grammar!
data cube provider, the provider for tabular data does not just allow selecting a subset of the
data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:

@@

The example works with a CSV file that records individual medals awarded in Olympic games.
The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
calculates total number of gold and silver medals for each team (country) before sorting the data.

When using the provider, the user specifies a sequence of operations. Members such as
\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
determine the operation type. Those are followed by operation parameters. For example, when grouping
data, we first select the key and then add a number of aggregations to calculate over the group.
Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
calculating the sum, average or the number of distinct values. As illustrated in
Section~\ref{sec:cases}, this still allows us to construct a wide range of practical queries.

\begin{figure}
\centering
\begin{subfigure}[b]{0.5\textwidth}
  \centering
  \includegraphics[scale=0.25]{figures/cubetp}
  \vspace{0.5em}
  \caption{The World Bank provider allows users to first choose
    a country or a year and then an indicator.}
  \label{fig:cubetp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[scale=0.28]{figures/graphtp}
  \caption{The user specifies a path through the data, possibly with
    placeholders to select multiple nodes.}
  \label{fig:graphtp}
\end{subfigure}
\vspace{-0.5em}
\caption{Design of type providers for exploring data cubes and graph databases.}
\label{fig:three graphs}
\end{figure}

\subsubsection*{Graph Database Type Provider}
Our third type provider allows users explore data from graph databases. A graph database
consists of nodes representing entities and relationships between them. In the context of
data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.

The following example explores a database of Doctor Who characters and episodes. It retrieves
all enemies of the Doctor that appear in the Day of the Moon episode:

@@

We start from the \ikvd{Doctor} node and then follow two relationships. We use
\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
to follow links to all enemies of the Doctor and then specify
\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
to only select only such enemies that appear in a specific episode. The resulting query
is illustrated in Figure~\ref{fig:graphtp}.

The provider works with any graph database and generates members automatically, based on the
data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
relationships. The results returned by the provider is a table of properties of all nodes
along the specified path. As discussed in the next section, such table can be further queried
using the tabular data type provider.

\newpage
% ==================================================================================================

\section{Motivation}
\label{sec:motivation}

The Gamma aims to adapt the recent innovations in programming language research, especially the
work on type providers, into a form where it could be used in practice by journalists and other
non-expert interested in data exploration. We start with a careful consideration of our target
application domain, i.e.~data analyses produced by journalists and citizen data scientists that
are published online. We look at both practical requirements for such programming environment
and requirements arising from our focus on journalism. This analysis is based on the author's
experience of collaborating with journalists\footnote{Citations removed to preserve anonymity.},
review of literature on data journalism, e.g.~\cite{ddj,edcj17,edcj18} and more general trends in
journalism.

\subsection{Open Journalism}
Journalism continually develops and responds to the many challenges it faces \cite{future}.
Two recent challenges are relevant to our work. The first is building trust in media.
One way of establishing trust in the age of fake news is to be more transparent about editorial
decisions, process and original sources.

@@

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/nyt}
\caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
(engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
\label{fig:nyt}
\end{figure}

\subsection{End-user Data Exploration}
Our aim is to make programmatic data exploration accessible to journalists, but we want to keep
the desirable properties of text-based programming. In particular, source code of a data
exploration should provide a full reproducible record of how the data analysis has been done.
As end-users, journalists have a number of interesting characteristics. They work under tight
deadline and data exploration is only a complementary skill. They also need to work with a wide
range of data sources, including big data tables (e.g.~Iraq War documents leak) or graph
databases (e.g.~Panama Papers). This leads to a number of practical requirements on the programming
environment.

@@

\paragraph{Learning without Experts}
Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
are unlikely to be available, so the tool should support learning from examples. When looking at
a work done and published by another person, the user needs to see (and be able to understand)
how a task was completed.

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/ft}
\caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
\label{fig:ft}
\end{figure}

We identified a number of requirements for a data exploration tool for journalists earlier.
First, using the tool should result in transparent and reproducible data analyses that can be
understood and modified by non-experts. Second, the tool should minimze the number of concepts
that the user needs to understand and, subsequently, allow them to learn through exploration
and from examples. In this section, we discuss how our design aims to satisfy those requirements.

\subsection{Lowering the Barrier to Entry}
Data exploration tasks, such as querying tables, have a certain irreducible complexity.
Regardless of the interface, the user will be exposed to concepts such as filtering,
sorting and grouping. Our design aims to lower the barrier to entry by stratifying the concepts
into first-level \emph{iterative prompting} principle and second-level \emph{domain specific
languages} for each kind of data source.
The user needs to master the iterative prompting principle in order to start working
with the system. However, using individual data sources can be mastered gradually.

\paragraph{Iterative Prompting}
Iterative prompting is an interaction principle where the user repeatedly invokes an
auto-complete prompt and makes a selection from the offered options. The principle is closely
related to both code completion and the use of command line. Iterative prompting is novel as an
overarching interaction principle for program construction.

@@

\subsection{Building the Magic Escalator of Knowledge}
As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
analyse data as their secondary task. The Gamma aims to support such users by having a low
barrier for entry and making it easy to learn independently.

@@

>>>>

\begin{figure}
\includegraphics[width=.6\columnwidth]{figures/thegamma-annot}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\end{figure}

@@

\begin{itemize}
\item We introduce the iterative prompting principle in The Gamma
  and show that it can be used for querying of distinct data sources including data tables, graph
  databases and data cubes (Section~\ref{sec:overview}).

% TODO REWRITE

@@

\subsection{Type Providers for Data Querying}
\label{sec:overview-tps}

The Gamma can be extended to support any kind of data source by implementing a \emph{type provider}.
Conceptually, a type provider defines a domain-specific language for exploring data of a particular
kind. Techincally, it generates object types with members (such as \ikvd{sort by} or \ikvd{Count is})
that are accessed via iterative prompting. We describe type provider for exploring data cubes
(inspired by an example from Syme et al. \cite{inforich}), tabular data (based on theory developed
by Petricek \cite{dotdriven}), and a novel type provider for exploring graph databases.

\subsubsection*{Data Cube Type Provider}
Our first type provider allows users to explore data form a data cube. For example, the World
Bank\footnote{The data is retrived using an API available at: \url{https://data.worldbank.org/}}
collects a range of indicators about many countries in the world each year. The data set is a
three-dimensional cube with dimensions corresponding to countries, indicators and years. The
following example uses the provider to access CO$_{2}$ emission data for United States:

\begin{thegamma}
worldbank.byCountry.'United States'.'Climate Change'.'CO2 emissions (kt)'
\end{thegamma}

\noindent
As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
from the data cube. Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
restricts the cube to a two-dimensional plane. We then choose an indicator category
and a specific indicator \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining
a time series with years as keys and emission data as values. Similarly, we could first filter the
data by a year or an indicator. The same mechanism can be used for exploring the UK government
expenditure:

@@

\noindent
The dimensions of the cube are government services, years and value type (adjusted, nominal,
per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type.
As there is no widely used standard format for data cubes, adding another data cube to The Gamma
currently requires implementing a new type provider.

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
the example from Figure~\ref{fig:thegamma}:

@@

\noindent
The example works with a CSV file that records individual medals awarded in Olympic games.
The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
calculates total number of gold and silver medals for each team (country) before sorting the data.

When using the provider, the user specifies a sequence of operations. Members such as
\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
determine the operation type. Those are followed by operation parameters. For example, when grouping
data, we first select the key and then choose a number of aggregations to calculate over the group.
Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
calculating the sum, average or the number of distinct values. As illustrated in
Section~\ref{sec:cases}, this is sufficient to construct a wide range of practical queries.

\begin{figure}
\centering
\begin{subfigure}[b]{0.5\textwidth}
  \centering
  \includegraphics[scale=0.25]{figures/cubetp}
  \vspace{0.5em}
  \caption{Exploring World Bank data using the data cube type provider, users
    choose values from two dimensions to obtain a data series.}
  \label{fig:cubetp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[scale=0.28]{figures/graphtp}
  \caption{To query graph data, the user specifies a path through the data, possibly with
    placeholders to select multiple nodes.}
  \label{fig:graphtp}
\end{subfigure}
\vspace{-0.5em}
\caption{Design of type providers for exploring data cubes and graph databases.}
\label{fig:three graphs}
\end{figure}

\subsubsection*{Graph Database Type Provider}
Our third type provider allows users to explore data from graph databases, which store
nodes representing entities and relationships between them. In the context of data journalism,
graph databases have been used for example in The Panama Papers reporting \cite{panama}.
The following example explores a database of Doctor Who characters and episodes. It retrieves
all enemies of the Doctor that appear in the Day of the Moon episode:

@@

\noindent
We start from the \ikvd{Doctor} node and then follow two relationships. We use
\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
to follow links to all enemies of the Doctor and then specify
\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
to select only enemies that appear in a specific episode. The resulting query
is illustrated in Figure~\ref{fig:graphtp}.

The provider works with any graph database and generates members automatically, based on the
data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
relationships. The results returned by the provider is a table of properties of all nodes
along the specified path. As illustrated by an example discussed in Section~\ref{sec:cases},
the returned table can be further queried using the tabular data type provider.

\subsection{Journalism}

We start with a careful consideration of our target
application domain, i.e.~data analyses produced by journalists and citizen data scientists that
are published online. We look at both practical requirements for such programming environment
and requirements arising from our focus on journalism. This analysis is based on the author's
experience of collaborating with journalists\footnote{Citations removed to preserve anonymity.},
review of literature on data journalism, e.g.~\cite{ddj,edcj17,edcj18} and more general trends in
journalism.

% Open Journalism

Journalism continually develops and responds to the many challenges it faces \cite{future}.
Two recent challenges are relevant to our work. The first is building trust in media.
One way of establishing trust in the age of fake news is to be more transparent about editorial
decisions, process and original sources.

@@

\subsection{Lowering barriers}
Our aim is to make programmatic data exploration accessible to journalists, but we want to keep
the desirable properties of text-based programming. In particular, source code of a data
exploration should provide a full reproducible record of how the data analysis has been done.
As end-users, journalists have a number of interesting characteristics. They work under tight
deadline and data exploration is only a complementary skill. They also need to work with a wide
range of data sources, including big data tables (e.g.~Iraq War documents leak) or graph
databases (e.g.~Panama Papers). This leads to a number of practical requirements on the programming
environment.

@@

Data exploration tasks, such as querying tables, have a certain irreducible complexity.
Regardless of the interface, the user will be exposed to concepts such as filtering,
sorting and grouping. Our design aims to lower the barrier to entry by stratifying the concepts
into first-level \emph{iterative prompting} principle and second-level \emph{domain specific
languages} for each kind of data source.
The user needs to master the iterative prompting principle in order to start working
with the system. However, using individual data sources can be mastered gradually.

\paragraph{Iterative Prompting}
%Iterative prompting is an interaction principle where the user repeatedly invokes an
%auto-complete prompt and makes a selection from the offered options. The principle is closely
%related to both code completion and the use of command line. Iterative prompting is novel as an
%overarching interaction principle for program construction.

@@

\subsection{Learnign without experts}

Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
are unlikely to be available, so the tool should support learning from examples. When looking at
a work done and published by another person, the user needs to see (and be able to understand)
how a task was completed.

As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
analyse data as their secondary task. The Gamma aims to support such users by having a low
barrier for entry and making it easy to learn independently.

@@

\newpage

% \begin{figure}
% \includegraphics[width=1\columnwidth]{figures/nyt}
% \caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
% (engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
% \label{fig:nyt}
% \end{figure}

% \begin{figure}
% \includegraphics[width=1\columnwidth]{figures/ft}
% \caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
% on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
% \label{fig:ft}
% \end{figure}

% ==================================================================================================
~
\newpage
